\LARGE \bf Stochastic Uncertainty Propagation in Power System Dynamics using Measure-valued Proximal Recursions

### 2022-03-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

{\nib{ \blue \ul{Our responses to the sub-comments below}.

(1) \textbf{Regarding usage of classical model:} Whether a model is relevant depends on the problem context. As pointed out in the last paragraph of Sec. II-B, the current state-of-the-art in power system literature for the problem we are addressing is ref. [45] with second order dynamics + single machine infinite bus ($n=1$ in our manuscript's notation). Given this context, our manuscript enables a new kind of computation that can handle moderately large $n$ with second order sample path dynamics. This computation is online and approximates neither the statistics nor the dynamics.  \ul{We clarify two important points here}:\\ \textbf{[1a]} The proposed proximal recursion framework on the space of probability measures is not conceptually limited by the second order dynamics, and we expect follow up research to design custom proximal recursions for \ul{higher order synchronous machine models}. In this sense, the present paper should be seen as the stepping stone laying the foundation for handling more complex dynamical models. \ul{The authors have been completely up front by explicitly mentioning this in the Conclusions section}. \\ \textbf{[1b]} Deriving proximal recursion requires understanding the generalized gradient flow structure over probability measures. \ul{Our research disposition is that one cannot bypass understanding the generalized gradient flow structure for the simpler second order sample path dynamics before attempting the same for more complex sample path models}.

(2) \textbf{Regarding randomness in active and reactive power:} Please note that the randomness mentioned by the reviewer is not ignored. In the simulation, their effects are aggregated as parametric uncertainty in the randomness of $P_{i}$. Also notice that to say this parametric uncertainty in $P_i$ manifests in the Kron reduced form, is equivalent to assuming first order stochastic dynamic uncertainties in the unreduced load dynamics. If one has an explicit (possibly higher order) stochastic load dynamics model, then the proposed framework applies as is by first computing the law of the stochastic process $P_{i}$ in the network reduced model, and then applying the proximal computation.

@@

\textbf{Regarding OU process in load:} \ul{The reviewer might have missed the following key point of our paper}. The Kron reduced second order stochastic model implies that in the unreduced sense, the load dynamics was first order noisy Kuramoto, in addition to the synchronous machine dynamics being second order noisy Kuramoto. Therefore, we do have stochastic load dynamics, which is in fact more realistic than OU process model. While OU process assumes linear drift with constant diffusion, whereas our setting considers the load SDEs having drifts with trigonometric nonlinearity and constant diffusion.

}}

\noindent
{\bf R1.4. Comments by reviewer}:\\
{\em There is no comparison of the proposed method compared with other state-of-the-art methods. Therefore, it is hard for the readers to judge the importance of this article.}

{\nib{ \blue \ul{Regarding the comparison with existing methods}, we have provided comparison with Monte Carlo in the revised manuscript's Figs. 12, 13 and 14; \ul{among these Figs. 13 and 14 are new} (in response to comment \textbf{R4.7}). Also, the \ul{new Fig. 8 in the revised manuscript shows the consistency for the rotor angle stochastic predictions w.r.t. Monte Carlo for the simulation reported in Sec. V-A}, as requested by comment \textbf{R2.5}. Please also see our response to comment \textbf{R4.2} where we explain why propagating the joint PDF for even a modest 5 generator system with discretization-based PDE solvers such as finite element simulation is challenging with general purpose computers.}}

@@

(3) \textbf{Regarding depict nonparametric PDF}: \ul{This paper is not about nonparametric \emph{depiction} but about nonparametric \emph{computation} of the transient joint state PDFs}. For example, Monte Carlo followed by function approximation via Kernel density estimates and its variants, or other function approximants (e.g., RBF, Gaussian mixture, neural networks) are all parametric methods. We enable computing the value of joint state PDF at a time-varying sample location \emph{online (not as offline post-processing)} without assuming any functional representation of the transient joint which the computed values correspond to. Notice that we are doing this joint computation without gridding the state space, which would be impossible had we followed the traditional scientific computing route: finite element/volume methods, tensor decomposition techniques etc. \ul{We believe the proposed computational framework is novel, highly non-trivial and is specifically useful in this power system uncertainty propagation context}.
}}

\noindent
{\bf R1.6. Comments by reviewer}:\\
{\em The simulation time is too short to convince the readers. Only 1 second is considered here. Note that the errors can be quickly accumulated in the iteration.}

{\nib{ \blue Please note that our joint PDF computation has no truncation error accumulation issue (as in for example, moment based propagation techniques where such errors arise from a priori assuming finite dimensionality of the sufficient statistic). \ul{Furthermore, the new Fig. 10 reports the total computational time for 1 minute time horizon}. Please also see the additional error statistics included in the revised version's Fig. 13 and Fig. 14.}}

@@

Having said that, we believe our technical contribution is indeed interdisciplinary in the sense it is not mere translational research. In other words, we are not simply importing ideas from a different domain and applying them to a power system problem. Instead, we make novel contributions at the (possibly unanticipated) intersection of two areas: measure-valued gradient flows and power system dynamics. The presentation of the paper reflects the same, perhaps rightfully so. Notice in particular that our presentation is bottom-up: we start from power system literature review and mention that the problem addressed in our paper was already recognized as a technical challenge in that literature. We move onto the theoretical ideas and tools much later in the manuscript, as the build up demands.

We expect a significant share of the readers of this work to be power system engineers, but not exclusively so since this contribution will also be of interest to the researchers in control, optimization and applied mathematics. We do not believe this issue is exceptional for our paper since several high impact papers in the IEEE Trans. Power Systems were indeed at the intersection of these areas, and have positively contributed toward increasing interdisciplinary research collaboration.}}

\noindent
{\bf R2.3. Comments by reviewer}:\\
{\em How will the proposed approach be used for the cases in which dependencies among uncertainty sources exist? An explanation and illustration would be helpful.}

{\nib \blue{In our paper, we do not make any assumption on statistical independence of the uncertainties. \ul{In fact, all stochastic uncertainties are considered to be ``joint", i.e., statistically correlated}. Please note that the \ul{initial joint PDFs} (29) and (31) are taken to be in that form for the sake of concrete example simulations. Their only purpose, in the simulations reported in Sec V, is to generate joint samples, and to evaluate the initial joint PDF values at the joint samples thus generated. If this initial scattered point cloud data is supplied to us only as a numerical table, without specifying any functional form whatsoever, nothing changes in the proposed algorithm. Similarly, the parameters in Table I could instead be correlated random vectors, and all we need are the \ul{joint} parametric weighted point cloud data. In fact, this is how we expect the proposed computational framework to be applied in practice.}}

@@

\noindent
{\bf R2.5. Comments by reviewer}:\\
{\em With regard to the point above, this reviewer would expect to see uncertainty bounds for the system dynamic response such as the trajectory of the rotor angle. Without such information, the method's usability is difficult to gauge.}

{\nib \blue{Thanks for this comment. \ul{The revised manuscript's newly incorporated Fig. 8 shows the statistical uncertainty bounds for the rotor angle sample paths for the simulation in Sec. V-A}. The inline text citing this figure in Sec. V-A is highlighted in blue.}}

\noindent
{\bf R2.6. Comments by reviewer}:\\
{\em Instead of using a purely synthetic system, the authors are highly encouraged to choose available test cases such as the IEEE 118-bus system and a larger one, if possible.}

{\nib \blue{Please note that the synthetic system in Sec. V-B is actually simulating a much higher dimensional system than the IEEE 118-bus system where the 19 generators would have resulted in the transient joint PDFs with 38 dimensional supports. In comparison, the synthetic system is evolving joint state PDFs having 100 dimensional supports, i.e., in $\mathbb{T}^{50}\times\mathbb{R}^{50}$. Our only intention behind simulating the synthetic system was to show that the computation can handle large dimensional joint state PDF supports. We also point out that the parameter ranges reported in Table I are consistent with the existing literature: refs. [25, Sec. 5], [20], [56], [57]. We hope the reviewer agrees that this simulation is adequate for the intended purpose.}}

### 2022-03-09 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound1-TPWRS-01391-2021.tex

\noindent
{\bf R3.4. Comments by reviewer}:\\
{\em Figures 5 and 6 are unreadable.}

{\nib \blue{Thanks for this comment. In the previous version, the strict 10 page limit enforced small figure sizes. In the revised manuscript, \ul{Figures 5 and 6 are made larger}. Also, the Fig. 6 in the revised manuscript shows the transient $(\theta_i,\omega_i)$ bivariate marginals for \ul{all the five generators in Case II}; the previous version showed the same only for the generator 4 (bus 6). We believe these changes have improved clarity.

Regarding the style of plots, we had considered plotting the bivariate marginals as planar contour plots (as in Fig. 1 in this rebuttal document) \ul{but decided against it because of the following two reasons}.

\emph{First}, \ul{this kind of plot is scientifically misleading as they hide the fact that these bivariate marginal PDFs are not really supported on the plane, or the subsets thereof, but instead have cylindrical supports}. Our message that the proposed method explicitly accounts this manifold geometry in the computation, would be missed had we chosen to make the plots the other way.%as per populist cognitive bias on behalf of the readers. That the planar plots evoke a sense of familiarity cannot be a scientific argument in favor of it.

@@

{\nib \blue{We appreciate the positive remark.}}

\noindent
{\bf R4.2. Comments by reviewer}:\\
{\em The key contribution of this paper is that the proposed method can reduce the computation burden when solving for the transient PDF. However, no comparison is provided with existing methods. The authors should provide comparisons to the exiting approaches, for example, based on finite element discretization or other numerical methods. The test example could be performed using the 5-generator system or larger systems. Besides, referring to Figs. 8 and 9, the total computation time should be reported for a longer horizon.}

{\nib \blue{\ul{Regarding the comparison with existing methods}, we have provided comparison with Monte Carlo in the revised manuscript's Figs. 12, 13 and 14; \ul{among these Figs. 13 and 14 are new} (in response to comment \textbf{R4.7}). Also, the \ul{new Fig. 8 in the revised manuscript shows the consistency for the rotor angle stochastic predictions w.r.t. Monte Carlo for the simulation reported in Sec. V-A}, as requested by comment \textbf{R2.5}.

\ul{Please note that performing finite element simulation for 5-generator system is a very tall order since the state space then is 10 dimensional, and even if we do a very coarse 10 discretizations per dimension, then such a simulation would require us to evolve a joint PDF over $10^{10}$ grid points}. This requires dedicated high-performance computing architecture and an undertaking well beyond the scope of this paper. This also explains why the state-of-the-art for direct computation of transient joint PDF in power systems literature remains single-machine-infinite-bus $(n=1)$ as in the 2013 ref. [45] which indeed performed the finite element simulation with a two dimensional discretization, and mentioned the difficulty of scaling this computational approach for even moderately large sized systems.

### 2022-08-12 Papers/Stochastic_Uncertainty_Propagation_TPS/RebuttalRound2-TPWRS-01391-2021R1.tex

\documentclass[12pt,onecolumn]{IEEEtran}
\usepackage{amsmath ,amssymb,euscript ,yfonts,psfrag,latexsym,dsfont,graphicx,bbm,color,amstext,wasysym,subfig,flushend,parskip}
\usepackage{bm,cite,soul,url}
\usepackage{verbatim}
\usepackage[normalem]{ulem}
\usepackage{amsthm}
%\usepackage{refcheck}
%\usepackage{showkeys}
%\usepackage{epstopdf,hyperref,pdfsync,url}
\graphicspath{{./},{./figures/}}
\newcommand{\mR}{{\mathbb R}}
\newcommand{\D}{{\mathbb D}}
\newcommand{\E}{{\mathbb E}}
\newcommand{\mcN}{{\mathcal N}}
\newcommand{\mcR}{{\mathcal R}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\tr}{\operatorname{trace}}
\newcommand{\ignore}[1]{}
\newcommand{\mike}{\color{magenta}}

\newcommand{\blue}{\color{blue}}
\newcommand{\red}{\textcolor{red}}

\newcommand{\bQ}{{\bm{Q}}}
\newcommand{\bPi}{\bm{\Pi}}
\newcommand{\bv}{{\bm{v}}}
\newcommand{\balpha}{{\bm{\alpha}}}
\newcommand{\bbeta}{{\bm{\beta}}}
\newcommand{\bq}{{\bm{q}}}
\newcommand{\bp}{{\bm{p}}}
\newcommand{\bbf}{{\bm{f}}}
\newcommand{\differential}{{\mathrm{d}}}

\newtheorem{claim}{Claim}

\def\spacingset#1{\def\baselinestretch{#1}\small\normalsize}
\setlength{\parskip}{10pt}
\setlength{\parindent}{20pt}
\spacingset{1}

\definecolor{grey}{rgb}{0.6,0.6,0.6}
\definecolor{lg}{rgb}{0.6,.6,0.6}
\newcommand{\nib}{\noindent  {\bf Response:} }

\begin{document}
%\newcommand{\bp}{{\bm{p}}}
\noindent
{\large {\bf Revision of
Manuscript Number}: TPWRS-01391-2021.R1\hfill \today\\[.1in]
{\bf Journal}: IEEE Transactions on Power Systems \\[.1in]
{\bf Author}: Abhishek Halder, Kenneth F. Caluya, Pegah Ojaghi, Xinbo Geng}\\

\section*{\large \bf Response to decision:}

{\noindent\blue
The perceptive input from all reviewers and the AE are much appreciated.

Please find below itemized responses to the individual comments addressing the concerns and detailing the revisions. All the edits in the revised manuscript are marked in blue.\\

\noindent
Sincerely,\\
The authors
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\newpage
\spacingset{1}

\section*{\large \bf Response to AE:}

\noindent
{\bf AE.1. Comments by AE}:\\
{\em Authors are requested to revise and resubmit the manuscript after addressing all reviewer comments that persist.
}

{\nib {\blue We. 
}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\spacingset{1}

\section*{\large \bf Response to Reviewer \#1:}

\noindent
{\bf R1.1. Comment by reviewer}:\\
{\em No further comments.}

{\nib {\blue Thank you for your time and efforts to help improve the manuscript.}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{\large \bf Response to Reviewer \#2:}

\noindent
{\bf R2.1. Comments by reviewer}:\\
{\em This reviewer appreciates the authors' effort in revising the manuscript based on the feedback in the previous round while clarifying and addressing the remarks with due diligence. The revised version reads well and adequately takes into account several relevant facets. The authors are kindly requested to proofread the article prior to final submission.}

{\nib \blue{Thank you, we have thoroughly proofread the article. We appreciate your comments from previous rounds to help us improve the manuscript.}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{\large \bf Response to Reviewer \#3:}

\noindent
{\bf R3.1. Comments by reviewer}:\\
{\em First and very importantly, the authors are suggested to readdress my previous two questions.\\
\noindent(1) Regarding my previous question about the classical model, I need to reemphasize here that the model is too simplistic that does not sufficiently reflect the responses from the exciter and the governor. Therefore, for a full paper on IEEE transactions on power systems, this model is incomplete, and cannot reflect reality. Also, for such an over-simplistic model, I still doubt the scalability of the method. Unfortunately, this problem is not addressed and improved after the revision.\\
\noindent(2) Regarding my previous questions about the comparison studies with the state-of-the-art method, indeed, I am very disappointed to find that the revised version still has added no comparison study with the state-of-the-art method at all. The ONLY comparison is still with Monte Carlo.}

{\nib \blue{The 
}}

\noindent
{\bf R3.2. Comments by reviewer}:\\
{\em Second and most importantly, I disagree with the theory and the claim behind the SDE model used in this paper.1.     Here, I am very familiar with the Ito SDE model in power systems that you cited here in your refs [16], [17], [45].

Although the authors claim the model you used here in equation (7) is the same as the ones used in ref[16] and ref[17] and extended from ref[45], I would like to tell the authors: NO, you are not using the same one. 16][17][45] are all right, but the one used in this article is incorrect. Please pay very special attention to [16][17], they only add uncertainties directly on differential equations, namely wi here, but not from P and Q.  This is because transferring the power system model from DAE to Ito-SDE is still an unsolved problem where P and is contained in algebraic equations while Ito-SDE only considers differential equations. Also, [45] is one machine system that does not even have a network. That is different from your problem. Here, the authors might argue that you are using the Kron reductions to remove algebraic equations. However, I have to tell the authors the Kron reductions to apply to the SDE model here. It is because the Kron reduction is relying on constant P and Q to derive a constant admittance matrix.  However, the P and Q in this manuscript are also assumed to be uncertain, and also claimed in the response letter to be time-varying stochastic processes. Then, its assumptions do not hold.  Then, its Ito SDE form is inconsistent.  I highly suggest to the authors study the following article carefully and learn more about the basics of Kron reduction.

Time-domain generalization of Kron Reduction, MK Singh, S Dhople, F Dorfler, GB Giannakis.
}

{\nib \blue{\ul{The.}}
}

\noindent
{\bf R3.3. Comments by reviewer}:\\
{\em The simulation results are weird. Just a simple check at fig 8. How a power system dynamic simulation with line 13 tripped can has no oscillations at all within the first 1 second?  If so, it has no practical meaning at all.}

{\nib \blue{The.}}

\noindent
{\bf R3.4. Comments by reviewer}:\\
{\em Third, my engineering motivation for me is not clear. The authors reclaim multiple times about the joint pdf for the output system state.  I understand that we consider joint pdf in the model inputs and understand its practical meaning. However, among the current literature, to my best knowledge, I saw no previous efforts in obtaining the ``joint" pdf of outputs. They only seed for marginal pdf for output states. So an important question is what motivates us to get joint pdf, instead of individual pdf, for model output? For which exactly applications, you must know the joint pdf of the output states?  What is its practical meaning and why it is important?}

{\nib \blue{Thanks for this comment. The}}

\noindent
{\bf R3.5. Comments by reviewer}:\\
{\em Also, the authors should put a very clear position about some of the references that you are making use to argue with the reviewer. For example,  [45] is a 10-year ago paper. Why do you claim it to the state-of-the-art?  How can this argument convince the reviewer?}

{\nib \blue{The }}

\noindent
{\bf R3.5. Comments by reviewer}:\\
{\em Regarding Ref [23]F. Milano, and R. Zarate-Minano, ``A systematic method to model power systems as stochastic differential-algebraic equations", IEEE transactions on Power Systems, Vol. 28, No. 4, pp. 4537--4544, 2013. I mentioned this article in the first-round review, not for the purpose of a simple cite, but trying to help you to improve the model and make your work realistic. However, a suggestion is ignored. This is indeed a pity. Besides, please do not argue again that this is a different work because this ref is well-known in modeling realistic stochastic power system dynamics, which should be the foundation of your model.}

{\nib \blue{The }}

\noindent
{\bf R3.6. Comments by reviewer}:\\
{\em In general, I need to claim here that math is fine in itself, but only publishable when connected to or interpreted correctly in a convincing engineering application.}

{\nib \blue{The }}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{\large \bf Response to Reviewer \#4:}

\noindent
{\bf R4.1. Comments by reviewer}:\\
{\em The authors have successfully addressed most of the reviewer's previous concerns. However, the reviewer still has some questions, mostly regarding the numerical simulations.}

{\nib \blue{Thank you. Please find our itemized responses to the questions related to numerical simulations.}}

\noindent
{\bf R4.2. Comments by reviewer}:\\
{\em It is stated that ``The Fig 9 shows that the computational time needed to perform the proximal updates, for all $t_k$, are indeed orders-of-magnitude faster than $h$."
This statement doesn't seem to hold, since $h=10^{-3}$ and the computational time lies within the range of around $10^{-3}$ to $10^{-2}$. How does the computational time needed to perform the proximal updates relate to time step $h$? If they are not strongly related, does that mean the step size $h$ is constrained to be greater than that computational time?}

{\nib \blue{The 
}}

\noindent
{\bf R4.3. Comments by reviewer}:\\
{\em In Fig. 10, could the authors please explain why the computational time of Case I is less than that of Case II?
The computational cost for the 5-generator system is roughly 4 times that of Case II. Could the authors comment if such a scale makes sense intuitively considering the state dimension of the two systems?}

{\nib \blue{Thanks for this comment. The ...}}

\noindent
{\bf R4.4. Comments by reviewer}:\\
{\em Both Fig. 12 and Fig. 13 demonstrate a decreasing trend as k increases. Could the authors comment on this and link this to the property of the proposed proximal algorithm? Similarly, why is the trend in Fig. 14? Comparing Figs. 12, 13 and Fig. 14, did the authors omit the initial time instant in Figs. 12, 13?}

{\nib \blue{Thanks for ...}}

\end{document}
